High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-04-24 notebooks/mnist.qmd
paper updates

@@

```{julia}
# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(

    Dense(n_in, n_hidden, activation),
    # Dense(n_hidden, n_hidden, activation),
    # Dense(n_hidden, n_hidden, activation),

    # Dense(n_in, n_hidden),
    # BatchNorm(n_hidden, activation),
    # Dense(n_hidden, n_hidden),
    # BatchNorm(n_hidden, activation),

    Dense(n_hidden, n_out),
)
# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.1, σ=activation)
# builder = MLJFlux.MLP(
#     hidden=(
#         n_hidden,
#         n_hidden,
#         n_hidden,
#     ), 
#     σ=activation
# )
α = [1.0,1.0,5e-3]

# Simple MLP:
mlp = NeuralNetworkClassifier(
    builder=builder, 
    epochs=epochs,
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
)

# Deep Ensemble:
mlp_ens = EnsembleModel(model=mlp, n=5)

# Joint Energy Model:
𝒟x = Uniform(-1,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=1
)
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=α,verbosity=10,
        # use_gen_loss=false,
        # use_reg_loss=false,
    ),
    sampling_steps=20,
    epochs=epochs,
)

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=5)
```

```{julia}
cov = .95
conf_model = conformal_model(jem_ens; method=:adaptive_inductive, coverage=cov)
mach = machine(conf_model, X, labels)
fit!(mach)
M = ECCCo.ConformalModel(mach.model, mach.fitresult)
```

>>>>

## Anecdotal Evidence

### Examples in Introduction

#### Wachter and JSMA

```{julia}
# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.9
T = 50
```

```{julia}
# Search:
opt = Flux.Optimise.Adam(0.01)
generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
```

#### REVISE

```{julia}
using CounterfactualExplanations.Models: load_mnist_vae
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
```

```{julia}
# Define generator:
generator = REVISEGenerator(
  opt = opt,
  λ=0.01
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x, target, counterfactual_data, M, generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
```

### ECCo

@@

```{julia}
# Data:
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(-1,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=1,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

```{julia}
# Simple MLP:
mlp = NeuralNetworkClassifier(
    builder=builder, 
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
)

# Deep Ensemble:
mlp_ens = EnsembleModel(model=mlp, n=n_ens)

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=20,
)

# JEM with adversarial training:
jem_adv = deepcopy(jem)
# jem_adv.adv_training = true

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=n_ens)

# Deep Ensemble of Joint Energy Models with adversarial training:
# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)
```

```{julia}
# Train models:
function _train(model, X=X, y=labels; cov=.95, method=:simple_inductive)
    conf_model = conformal_model(jem_ens; method=method, coverage=cov)
    mach = machine(conf_model, X, y)
    fit!(mach)
    M = ECCCo.ConformalModel(mach.model, mach.fitresult)
    return M
end
model_dict = Dict(mod_name => _train(mod) for (mod_name, mod) in models)
```
